Extended Abstract

### 2021-07-17 paper/paper.tex
Apply some of the minor changes suggested in reviews

\newcommand{\subst}[2]{#1 ~ \func{\triangleleft} ~ #2}
\newcommand{\tSubst}[2]{\type{Subst}~#1~#2}
\newcommand{\interpr}[1]{\func{[\![} #1 \func{]\!]}}

@@

To type check a \picalc{} process with shared and linear types we must assign a type to every communication channel created within the process.
To do so we can either ask the user for type annotations \cite{ZalakainD21}, or we can infer types by looking at how channels are used.
To reduce the burden on the user, we follow the latter approach: we traverse processes bottom-up while keeping a typing context with \emph{metavariables} (\emph{holes}) and collecting typing constraints on metavariables, in line with co-contextual typing \cite{ErdwegBKKM15}.
Keeping a strictly bottom-up information flow has the additional advantage of making typing inference easily parallelisable.
Constraints with a most general solution are solved into substitutions and applied to the typing context, constraints without one must be kept around for later: applying them as substitutions risks over-constraining the problem down the line.
Armed with a typing inference algorithm that for a process $P$ infers the most general typing context $\Gamma$ and some typing constraints, type checking that $\Delta \vdash P$ for some $\Delta$ amounts to emitting the extra constraint $\Gamma = \Delta$, assuming the metavariables in $\Gamma$ and $\Delta$ are disjoint.

@@

Both types and usage annotations contain metavariables.
We use a common set of metavariables for both --- this makes their handling uniform.
A context of kinds \(\gamma\) keeps track of whether a metavariable is of type kind $\ktype$ or usage annotation kind $\kusage$.
We refer to a metavariable of kind \(k\) in a kinding context \(\gamma\) as $\tvar{\gamma}{k}$.
We can now define usage annotations and types in one go (we define $\tzero$, $\tone$ $\tomega$ at once for brevity):
\begin{mathpar}
  \inferrule {m : \tvar{\gamma}{k}} {\tmvar~m : \tkind{\gamma}{k}}

@@

  \inferrule { }
  {\tunit : \tkind{\gamma}{\ktype}}

@@

  \inferrule {s : \tkind{\gamma}{\ktype} \\ t : \tkind{\gamma}{\ktype} }
  {\tsum~s~t : \tkind{\gamma}{\ktype}}

@@

\paragraph{Context Splits}
\label{types}
A context $\Gamma$ of type $\tCtx{n}{\gamma}$ is a list of $\ttype{\gamma}$ of size $n$.
We define context splits $\tSplit{\Gamma}{\Delta}{\Theta}$ pointwise on types.
Splits on types are defined recursively on usage annotations --- note that only the usage annotations at the top of a channel are split, not those within its payload.
Splits on usage annotations are defined as follows:
\begin{mathpar}
  \inferrule { } {\tSplit{x}{x}{\tzero}}

@@

  \inferrule { } {\tSplit{\tomega}{x}{y}}
\end{mathpar}
Following \cite{Padovani15}, a context $\Gamma$ is unrestricted (non-linear) $\tun{\Gamma}$ if it can be split into itself: $\tSplit{\Gamma}{\Gamma}{\Gamma}$ (respectively $\tun{x}$ and $\tun{t}$ for usage annotations $x$ and types $t$).

@@

A kind-preserving substitution $\tSubst{\gamma}{\delta}$ maps usage annotations and type metavariables in $\gamma$ to usage annotations and types in $\delta$, that is, $\forall k \to \tvar{\gamma}{k} \to \tkind{\delta}{k}$.
The function $\subst{\sigma}{t}$ of type $\tSubst{\gamma}{\delta} \to (\forall k \to \tkind{\gamma}{k} \to \tkind{\delta}{k})$ performs the substitution by replacing all the metavariables in $t$ with their corresponding terms in $\sigma$.
Substitutions on constraints are defined pointwise on their arguments.

@@

Typing inference takes a process with $n$ free variables and returns a metavariable context $\gamma$, a typing context with $n$ free variables containing metavariables in $\gamma$, and a list of constraints on metavariables $\gamma$.
We define a similar function for typing inference on expressions, this time also returning a type $\tau$ of type $\ttype{\gamma}$:
\begin{flalign*}
& \func{inferProc} : \sProc{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} && \\
& \func{inferExpr} : \sExpr{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} \times \ttype{\gamma} &&
\end{flalign*}
These functions are total: if a process or expression is untypable its constraints will be unsolvable.

@@

For the expression $\scase~e~p~q$ we
  \begin{enumerate*}[label=\textcolor{gray}{\arabic*)}]
  \item infer $p$ resulting in $\theta_p, (\Theta_p, s) , Cs_p$;
  \item infer $q$ resulting in $\theta_q, (\Theta_q, t) , Cs_q$;
  \item infer $e$ resulting in $\delta_e, \Delta_e, Cs_e, \tau$;
  \item create a new metavariable context $\gamma$ and a new typing context $\Gamma$ where every type is a fresh metavariable in $\gamma$;
  \item take the union of the metavariables in $\gamma$, $\theta_p$, $\theta_q$ and $\delta_e$;
  \item take the union of the constraints in $Cs_p$, $Cs_q$ and $Cs_e$, and create new constraints $\eqconstr{\Theta_p}{\Theta_q}$, $\sumconstr{\Gamma}{\Delta_e}{\Theta_p}$, and $\eqconstr{\tau}{\tsum~s~t}$; and
  \item return $\Gamma$ as the inferred context.
\end{enumerate*}

>>>>

\newcommand{\subst}[2]{#1 ~ \func{\triangleleft} ~ #2}
\newcommand{\tSubst}[2]{\type{Subst}~#1~#2}
\newcommand{\interpr}[1]{\func{\llbracket} #1 \func{\rrbracket}}

@@

To type check a \picalc{} process with shared and linear types we must assign a type to every communication channel created within the process.
To do so we can either ask the user for type annotations \cite{ZalakainD21}, or we can infer types by looking at how channels are used.
To reduce the burden on the user, we follow the latter approach: we traverse processes bottom-up while keeping a typing context with \emph{metavariables} (\emph{holes}) and collecting typing constraints on metavariables, in line with co-contextual typing \cite{ErdwegBKKM15}.
Keeping a strictly bottom-up information flow has the additional advantage of making typing inference easily parallelisable.
Constraints with a most general solution are solved into substitutions and applied to the typing context, constraints without must be kept around for later: applying them as substitutions risks over-constraining the problem down the line.
Armed with a typing inference algorithm that for a process $P$ infers the most general typing context $\Gamma$ and some typing constraints, type checking that $\Delta \vdash P$ for some $\Delta$ amounts to emitting the extra constraint $\Gamma = \Delta$, assuming the metavariables in $\Gamma$ and $\Delta$ are disjoint.

@@

Both types and usage annotations contain metavariables.
We use a common set of metavariables for both --- this makes their handling uniform.
A context of kinds \(\gamma\) keeps track of whether a metavariable is of type kind $\ktype$ or usage annotation kind $\kusage$.
We define $\tvar{\gamma}{k}$ as the set of variables of kind \(k\) in a kinding context \(\gamma\), and $\tkind{\gamma}{k}$ as the set of types \(\ktype\) and usage annotations \(\kusage\) (we omit unit and sum types and define $\tzero$, $\tone$, $\tomega$ at once for brevity):
\begin{mathpar}
  \inferrule {m : \tvar{\gamma}{k}} {\tmvar~m : \tkind{\gamma}{k}}

@@

@@

@@

\paragraph{Context Splits}
\label{types}
A context $\Gamma$ of type $\tCtx{n}{\gamma}$ is a list of $\ttype{\gamma}$ of size $n$.
We define context splits $\tSplit{\Gamma}{\Delta}{\Theta}$ pointwise on types.
Splits on types are defined pointwise on usage annotations (but are not applied to a channel's payload) and splits on usage annotations are defined as follows:
\begin{mathpar}
  \inferrule { } {\tSplit{x}{x}{\tzero}}

@@

  \inferrule { } {\tSplit{\tomega}{x}{y}}

  \inferrule {\tSplit{i_x}{i_y}{i_z} \\ \tSplit{o_x}{o_y}{o_z}} {\tSplit{\tchan~i_x~o_x~t}{\tchan~i_y~o_y~t}{\tchan~i_z~o_z~t}}
\end{mathpar}
Following \cite{Padovani15}, a context $\Gamma$ is unrestricted (non-linear) $\tun{\Gamma}$ if it can be split into itself: $\tSplit{\Gamma}{\Gamma}{\Gamma}$ (respectively $\tun{x}$ and $\tun{t}$ for usage annotations $x$ and types $t$).

@@

A kind-preserving substitution $\tSubst{\gamma}{\delta}$ maps usage annotations and type metavariables in $\gamma$ to usage annotations and types in $\delta$, that is, $\forall \{k\} \to \tvar{\gamma}{k} \to \tkind{\delta}{k}$ for an implicit $k$.
The function $\subst{\sigma}{t}$ of type $\tSubst{\gamma}{\delta} \to (\forall \{k\} \to \tkind{\gamma}{k} \to \tkind{\delta}{k})$ performs the substitution by replacing all the metavariables in $t$ with their corresponding terms in $\sigma$.
Substitutions on constraints are defined pointwise on their arguments.

@@

Typing inference takes a process with $n$ free variables and returns a metavariable context $\gamma$, a typing context with $n$ free variables containing metavariables in $\gamma$, and a list of constraints on metavariables $\gamma$.
We define a similar function for typing inference on expressions, this time also returning a type in $\ttype{\gamma}$:
\begin{flalign*}
& \func{inferProc} : \sProc{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} && \\
& \func{inferExpr} : \sExpr{n} \to ~ \exists \gamma. ~ \tCtx{n}{\gamma} \times \tConstrs{\gamma} \times \ttype{\gamma} &&
\end{flalign*}
These functions are total: if a process or expression is untypable its constraints will be unsolvable.

@@

For the expression $\scase~e~p~q$ we
  \begin{enumerate*}[label=\textcolor{gray}{\arabic*)}]
  \item infer $p$ resulting in $(\theta_p, \Theta_p, s , Cs_p)$;
  \item infer $q$ resulting in $(\theta_q, \Theta_q, t , Cs_q)$;
  \item infer $e$ resulting in $(\delta_e, \Delta_e, Cs_e, \tau)$;
  \item create a new metavariable context $\gamma$ and a new typing context $\Gamma$ where every type is a fresh metavariable in $\gamma$;
  \item take the union of the metavariables in $\gamma$, $\theta_p$, $\theta_q$ and $\delta_e$;
  \item take the union of the constraints in $Cs_p$, $Cs_q$ and $Cs_e$, and create new constraints $\eqconstr{\Theta_p}{\Theta_q}$, $\sumconstr{\Gamma}{\Delta_e}{\Theta_p}$, and $\eqconstr{\tau}{\tsum~s~t}$; and
  \item return $\Gamma$ as the inferred context.
\end{enumerate*}
